Waveform Inversion via Source Extension: Overview as of Q1 2019

### 2019-10-23 repro/papers/trans/paper.tex
update

\title{Waveform Inversion via Source Extension: Major Questions and Research Plan}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

@@

\section{Proposed Project}.

Amongst the source extension options described in the previous section, only the surface extension has all of these properties:
\begin{itemize}
\item requires only additional storage roughly equivalent to the data volume
\item does not require prior knowledge of the source wavelet, in fact recovers an estimate as a by-product
\item inner problem (in variable projection - estimation of the extended source) has a cheap approximate solution, via back-propagation
\item relation with traveltime inversion persists, regardless of ray geometry
\end{itemize}

Therefoire I propose that we begin with a time-domain implementation of the surface source extension.

\section{Detailed Description}
The simplest version of the surface source extension assumes acoustic wave physics, and that the physical source locations lie on the horizontal surface $z=z_s$. The coefficient vector $c$ is the pair $(\kappa,\beta)$ of bulk modulus and buoyancy, and the wave operator $L[c]$ is given by the linear acoustics system:
\begin{eqnarray}
\label{eqn:awe}
\frac{\partial p}{\partial t} & = & - \kappa (\nabla \cdot \bv +
w(t)\delta(\bx-\bx_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t<0\nonumber\\ 
\bv & = & 0 \mbox{ for } t<0 
\end{eqnarray}
The physical source is modeled as an isotropic point constitutive law defect. As mentioned earlier, it is simple to accommodate more complex source models.

The dynamical field is the 4-tuple $u=(p,\bv)$. Either the pressure $p$ or the particle velocity $\bv$ could be sampled (or both) - presume that the pressure is sampled, that is $Pu = \{p(\bx_r,t;\bx_s)\}$, where $\{\bx_r\}$ are the receiver positions.

Extended sources are spread over $z=z_s$ (\ref{eqn:mssur}). The annihilator is the source localization penalty, equation \ref{eqn:msste}.

Denote by $J_{\alpha}[g,c;d]$ the objective function defined in \ref{eqn:esi}:
\begin{equation}
\label{eqn:jdef}
J_{\alpha}[g,c;d] = \frac{1}{2}( \|PL[c]^{-1}g -d \|^2 + \alpha \|A[g]\|^2) 
\end{equation}
To compute the objective 
\begin{itemize}
\item solve the acoustic wave system \ref{eqn:awe} with $g$ as in \ref{eqn:mssur}
\item sample at receiver locations and compute mean-square difference with pressure data traces, take mean square
\item add scaled mean square of annihilator applied to the source
\end{itemize}

The variable projection method \cite[]{GolubPereyra:03} applied to the objective \ref{eqn:esi} approaches its minimization via a nested algorithm: first minimize \ref{eqn:esi} over the source $g$ for given coefficients $c$, thus making $g = g[c]$ a function of $c$ (inner minimization); then substitute $g[c]$ in $J$ and minimize the resulting function of $c$ (outer minimization). The optimal $g[c]$ solves the first order necessary condition for minimization of $J$ over $g$: $\nabla_g J_{\alpha}[g,c;d] = 0$. For linear annihilators such as that defined by \ref{eqn:mssur}, $J$ is quadratic in $g$, so the first order necessary condition is also sufficient, and is equivalent to the linear {\em normal equation}:
\begin{equation}
\label{eqn:normal}
(S[c]^TS[c] + \alpha A^TA) g = S[c]^T d
\end{equation}
where I have written $S[c]=PL[c]^{-1}$ for the operator that maps the (extended) source to the data traces. Substitution of $g[c]$ in $J$ results in the {\em reduced objective}
\begin{equation}
\label{eqn:jred}
\tilde{J}_{\alpha}[c;d] = J_{\alpha}[g[c],c;d].
\end{equation}
The gradient would appear to involve the derivative of the inner solution $g[c]$ with respect to $c$, computation of which implies solution of additional linear systems. However by fortunate accident, the chain rule and the normal equation \ref{eqn:normal}
conspire to eliminate all terms involving derivatives of $g[c]$. Consequently the direction derivative of $\tilde{J}_{\alpha}$ in the direction $\delta c$ is
\begin{equation}
\label{eqn:tildejderiv}
D_c\tilde{J}_{\alpha}[c;d]\delta c = \langle D(S[c]g)\delta c,S[c]g-d \rangle|_{g=g[c]}
\end{equation}
whence
\begin{equation}
\label{eqn:l2grad}
\nabla \tilde{J}_{\alpha}[c;d] = D_c(S[c]g)^T|_{g=g[c]}(S[c]g[c]-d)
\end{equation}
Let us field-strip the expression on the right. The rightmost factor is the data residual, which is a by-product of many methods for the solution of the normal equation \ref{eqn:normal}. $D_cS$ is the linearization or Born approximation of $S$, or perturbation with respect to $c$, with $g$ fixed:
\[
S[c+\delta c]g \approx S[c]g + D_c(S[c]g)\delta c
\]
That is, $D_c(S[c]g)$ is the object commonly called the Born modeling operator, with source wavefield $g$. Its adjoint or transpose is $D_c(S[c]g)^T$ - as is well known, this is the RTM operator (or rather a version of it), computed via the adjoint state method in the time domain. Note that the Born and RTM operators need to match, i.e. both be derived from the same underlying full waveform modeling operator (in this case acoustodynamics, but the same would apply for any other wave theory).

The upshot is that the computation of the gradient goes like this:
\begin{itemize}
\item solve the normal equation \ref{eqn:normal} for $g[c]$
\item compute (or retain) the data residual $S[c]g[c]-d$
\item reverse-time migrate the data residual, using $g[c]$ to generate the source wavefield in the adjoint state method - the output is the gradient of $\tilde{J}_{\alpha}$.
\end{itemize}

To make this algorithm practical and efficient, effective preconditioning needs to be applied to both inner and outer problems.

\subsection{Preconditioning the Outer Problem: Gradient Smoothing}
Conditioning the outer (velocity etc.) update has different meaning for transmission versus reflection problems. One interpretation of ``transmission'' is that no reflection occurs on any ray path connecting source and receiver, whereever these may be located. Is is practically implicit in the use of ``ray path'' in this statement that the material model is at least mostly smooth, or essentially localized in low spatial frequencies. Smoothness is also one way to interpret ``lack of reflections''. The classic work explaining the need for model smoothness in tomography is \cite{DelpratLailly:92}. Since the surface source extension method gets its ability to identify kinematically correct velocity from its relation to tomography, it also requires model smoothness, at least for transmission.

A natural method for enforcing smoothness is use of a norm in model space that penalizes lack of smoothness. This method is also properly called preconditioning, which consists in altering the choice of norm to enhance convergence. For pure transmission, the choice recommended by \cite[]{DelpratLailly:92} is a norm from the Sobolev family. These are weighted $L^2$ (RMS) norms with weight operators indexed by a real parameter $s$:
\begin{equation}
\label{eqn:sob}
\|u\|^2_s = \langle u, W_s u \rangle_{L^2},\,\,W_s = (I - \nabla^2)^s
\end{equation}
Choosing $s>0$ penalizes lack of smoothness. According to the Sobolev Embedding Theorem and the basic well-posedness theorem for ordinary differential equations (which governs the behaviour of the ray system), it is neccessary to choose $s > 2 + d/2$ in dimension $d$, in order that metrizing the velocity (amongst other things) with the $s$-norm will guarantee sufficient smoothness that ray theory is well-behaved, and the link with tomography honored. For both $d=2$ and $d=3$, $s=4$ would work.

The gradient of a function is defined with respect to the domain norm. Using the $s$-norm, the gradient is
\begin{equation}
\label{eqn:grads}
\nabla_s \tilde{J}_{\alpha}[c,d] = W_s^{-1}\nabla \tilde{J}_{\alpha}[c,d]
\end{equation}
If $s>0$ then $W_s^{-1}$ is a smoothing operator. It is best implemented in the Fourier domain.

The most natural way view the reflection problem is to allow non-smoothness in $c$. It is not clear at this point what would be the optimal choice of weight operator in this case, or even whether there is a choice of Hilbert space structure that would offer an easily computable gradient and development of short-scale structures to generate reflections. GH used very low frequency data (in frequency domain) and Tihonov regularization. His results seemed to depend heavily on the regularization weight. Of course we also do not know how to connect the reflection case to a tomographic principle, so have no guidance for how a model space norm should be selected. These are the main target questions of this project.

\subsection{Preconditioning the Inner Problem: Time Reversal}
%In contrast to the outer problem, a preconditioner for the inner problem is available that works regardless of medium smoothness, so for both transmission and reflection (the latter with some caveats). 
In contrast to the outer problem, the inner problem admits a conventional preconditioner, that is, a change of variables that in the spaces of source traces and data traces respectively that moves the singular values of the modeling operator closer to 1. This preconditioner also creates an approximate solution operator for the inner problem, of the form
\begin{equation}
\label{eqn:appinv}
S[c]^{\dagger} = W_m[c]^{-1}S[c]^TW_d[c],
\end{equation}
in which $W_m$ and $W_c$ are $c$-dependent positive definite symmetric operators that define weighted inner products in domain and range of $S[c]$, and their associated weighted norms:
\begin{eqnarray}
\label{eqn:norms}
\langle f,g \rangle_m & = & \langle f, W_mg\rangle; \|f\|_m^2 = \langle f,f\rangle_m \nonumber \\
\langle f,g \rangle_d & = & \langle f, W_d g\rangle; \|f\|_d^2 = \langle f,f \rangle_d
\end{eqnarray}
We have implicitly used the notations $\langle \cdot,\cdot \rangle, \|\cdot \|$ (without subscripts) to denote the (unweighted) $L^2$ norm, in whatever form is appropriate for the arguments. That is what these symbols mean in display \ref{eqn:norms}, and will continue to mean.

The approximate inverse $S[c]^{\dagger}$ defined by the relation \ref{eqn:appinv} is the transpose of $S[c]$ with respect to the inner products $\langle \cdot,\cdot \rangle_m,, \langle \cdot,\cdot \rangle_d$, so $S[c]$ is approximately unitary in the same sense. Consequently the singular spectrum (which is also dependent on choice of inner product) is close to that of the identity - or most of it is, as we shall explain - and the convergence of  a Krylov subspace method like Conjugate Gradient (CG) Iteration converge relatively rapidly.

The system that needs solving, of course, is the normal equation \ref{eqn:normal},  so it is the convergence of CG or the like applied to that system that is the important issue. The normal equation is the first order necessary condition for minimization of the objective
\begin{equation}
\label{eqn:jdefs}
J_{\alpha}[g,c;d] = \frac{1}{2}( \|S[c]g -d \|^2 + \alpha \|A[g]\|^2).
\end{equation}
over the choice of surface source $g$. The choice of norms appearing in \ref{eqn:jdefs} is actually arbitrary, and the norms appearing in the first and second terms on the right-hand side do not even have to be the same.

%First the transmission case - a sort of rotated crosswell configuration. Idealize the receiver set as a continuum plane $z=z_r$, with $z_r > z_s$ - not essential but simpler.

%Two key facts and a geometric assumption underly the construction of the time-reversal preconditioner: (i) a local relation between the surface source and the normal particle velocity, and (ii) the uniqueness of solutions of the Dirichlet boundary value problem for the wave equation. Together these facts allow us to reconstruct the pressure field, or the significant part of it, from the transmission data, and to approximate a distributed source responsible for a given set of data traces.

%The geometric assumption is that the pressure field is {\em outgoing} at the recording datum: that is, that it is the restriction to $z>z_s$ of a pressure field that is negligible in $z<z_r$ for large enough time. ``Negligible'', rather than vanishing, as the latter is strictly speaking impossible: this property is asymptotic in frequency, and assured by an assumption about the ray field associated to the acoustic system.

%NB: the definition of the forward map (and the data) should include a taper $\phi(\bx_r,\bx_s)$ masking out the shot gather $d_r(\cdot,\cdot;\bx_s)$ for each source position $\bx_s$. However I will not explicitly mention this operation. I will also ignore it in the numerical examples, but there will be a price to pay.

To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$. As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.

The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
\begin{eqnarray}
\label{eqn:awedata}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h \delta(z-z_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\end{eqnarray}
Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata}, defined by integration by parts. A more general setting for this type of system provides an advantageous framework within which to explain the properties of these weak solutions.

Let $\omega \subset \bR^d$ be a smooth orientable surface. Then the distribution $\delta_{\omega}$ is well-defined: for $u \in C_0^{\infty}(\bR^d)$, 
\[
\langle \delta_{\omega}, u \rangle = \int_{\omega} u dS
\]
The system \ref{eqn:awedata} is a special case of
\begin{eqnarray}
\label{eqn:aweomega}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h \delta_{\omega}, \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\end{eqnarray}
in which $h \in C_0^{\infty}(\omega)$.

\begin{thm}
\label{thm:bvs}
Suppose that $\kappa,\beta$ are smooth. Then the system \ref{eqn:aweomega} has a unique weak solution $(p,\bv)$. Denote by $[v_n] \in C^{\infty}(\omega)$ the jump of the normal component of $\bv$ across $\omega$:
\[
[v_n](\bx) = \lim_{\epsilon \rightarrow 0} (\bv(\bx + \epsilon {\bf n}(\bx))-\bv(\bx - \epsilon {\bf n}(\bx)))\cdot {\bf n}(\bx),\,\bx \in \omega.
\]
Then $h(\bx) = [v_n](\bx), \bx \in \omega$.

%Let $p,\bv$ be the weak solution of the system \ref{eqn:awedata}, and assume that $\kappa, \beta$ are smooth. Denote by $[v_z]|_{z=z_s}$ the jump of the z-component of particle velocity in the z-direction:
%\[
%[v_z](x,z) = v_z^+(x,z)-v_z^-(x,z),  v_z^{\pm}(x,z)=\lim_{z'\rightarrow z^{\pm}} v_z(x,z')
%\]
%Then $[v_z](x,z)=0$ unless $z=z_s$, and
%\[
%h(x) = [v_z](x,z_s).
%\]
\end{thm}
For the proof, see  Appendix B.

For the rest of this section, $\omega$ will be a coordinate hyperplane, defined by $z=$ constant, with its usual orientation, and I will write $[v_z] = [v_b]$.

To begin with, change notation slightly, and write $S[c]h_s$ rather than $S[c]g$ with $g = h_s \delta(z-z_s)$, as $z_s$ will be presumed ot be the same for all shots. 
%The addition of the subscript indicates that the sources lie on the plane $z=z_s$ and the receivers are on the plane $z=z_r$. As will be apparent shortly, a similar operator with source and receiver planes swapped will also be important, hence the need for the additional information in the notation.

The pressure and velocity fields occuring in the computation of $d_r=p|_{z=z_r}=S[c]_{s,r}h_s$ form the solution of
\begin{eqnarray}
\label{eqn:awedata1}
\frac{\partial p}{\partial t} & = & - \kappa( \nabla \cdot \bv +
h \delta(z-z_s)), \nonumber \\
\frac{\partial \bv}{\partial t} & = & - \beta \nabla p,\nonumber \\
p & =& 0 \mbox{ for } t \ll 0 \nonumber\\ 
\bv & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\end{eqnarray}
Actually $p,\bv$ form a {\em weak solution} of the system \ref{eqn:awedata1}, defined by integration by parts.
\begin{thm}
\label{thm:vz2}
If $\kappa,\beta$ are constant near the $z=z_s$, then $v_z^-=-v_z^+$ so $[v_z] = 2 v_n^+$. 
\end{thm}
(for proof see Appendix 2). The formulation is for 2D propagation, but a precisely similar fact is true for 3D - in that case, $x$ is replaced everywhere by the horizontal coordinates.

The import of Theorem \ref{thm:bvs} is that we can read off the source if we can infer the (particle) velocity field near the source surface, from knowledge of the data traces (values of the pressure at $z=z_r$). The key observation that enables this feat is a well-posedness theorem for the wave equation:

\begin{thm} \label{thm:bvp}
Suppose that $\omega \subset \bR^d$ is a smooth orientable surface, $\omega = \partial  \Omega^- = \partial \Omega^+$, where $\Omega^{\pm}$ are the open exterior/interior regions defined by the outward unit normal field of $\omega$, and let $d \in C^{\infty}_0(\omega \times \bR)$. Then there exist unique smooth solutions $p^{\pm},\bv^{\pm}$ of the initial/boundary value problems 
\begin{eqnarray}
\label{eqn:awebdry}
\frac{\partial p^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \bv^{\pm} \mbox{ in } \Omega^{\pm}, \nonumber \\
\frac{\partial \bv^{\pm}}{\partial t} & = & - \beta \nabla p^{\pm} \mbox{ in } \Omega^{\pm},\nonumber \\
p^{\pm} & =& 0 \mbox{ for } t \ll 0\nonumber\\ 
\bv^{\pm} & = & 0 \mbox{ for } t \ll 0 \nonumber\\
\lim_{\bx \rightarrow \omega^{\pm}}p & = & d 
\end{eqnarray}
\end{thm}

Putting the two theorems together, obtain
\begin{cor}\label{thm:bvps}
Suppose that $d_s$ is a sufficiently regular function, = 0 for $t<0$, and denote by $p^{\pm},\bv^{\pm}$ the half-space solutions provided by Theorem \ref{thm:bvp}. Define the vector fields $p,\bv$ in $\bR^4$ by
\begin{eqnarray}
\label{eqn:putemtogether} 
p(x,z) &=& p^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
p(x,z_s)&=&d_s(x)\\
\bv(x,z) &=& \bv^{\pm}(x,z) \mbox{ for } \pm(z-z_s) > 0\\
\end{eqnarray}
Then $p,\bv$ is a weak solution of the system \ref{eqn:awedata1} with $h=[v]_n$.
\end{cor}

Now invoke the {\em outgoing assumption}: $p,\bv$ is the restriction to $z>z_s$ of a solution vanishing in $z<z_r$ for large $t$. This condition needs to be interpreted asymptotically, that is, $p,\bv$ is relatively smooth (lower order in frequency) for $t \gg 0, z<z_r$. It holds if the rays carrying significant energy in $p,\bv$ intersect $z=z_r$ transversally, and do not return to $z<z_r$ in forward time. The examples below will satisfy this condition, because the wave velocity is constant in $z > z_r$.

Note that isotropic point sources do not produce fields outgoing in a half space, in the sense just described - simply because they radiate in all directions. However, for any specified finite time of propagation, it is possible to replace a point source by an anisotropic source that fits the point source data to within an arbitrarily small error, and does produce outgoing fields. In fact, the numerical examples below will construct such sources.

This assumption implies that for $t$ sufficiently large, $d_r = p|_{z=z_r}$ is negligible for $t \gg 0$, so $p,\bv$ differs in $z_s<z < z_r$ by a negligible error from the solution of $p^-,\bv^-$, part of the solution of the  backwards-in-time  initial value problem, similar to the system \ref{eqn:awebdry}:
\begin{eqnarray}
\label{eqn:awetr}
\frac{\partial \tilde{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \tilde{\bv}^{\pm}, \pm(z-z_r)>0 \nonumber \\
\frac{\partial \tilde{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \tilde{p}^{\pm},\nonumber \\
\tilde{p}^{\pm} & =& 0 \mbox{ for } t \gg 0,\nonumber\\ 
\tilde{\bv}{\pm} & = & 0 \mbox{ for } t \gg 0,\nonumber\\
\lim_{z \rightarrow z_r^{\pm}}\tilde{p}^{\pm} &=&  d_r, \nonumber\\,
\tilde{p} &=& \tilde{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
\tilde{\bv} &=& \tilde{\bv}^{\pm}, \pm(z-z_r)>0.
\end{eqnarray}

This must be so, as $p,\bv$ satisfies the wave equation system (first two conditions), and the rest as well except for a negligible error. The last condition must be interpreted consistently, that is $\tilde{p}=0$ for $t$ large enough that $d_r$ is negligible. (In practice consistency will be obtained by replacing $t \gg 0$ respectively $t \ll \infty$ by $t > t_{\rm max}$, $t < t_{\rm max}$ for a suitably large $t_{\rm max}$). Then the standard energy estimate shows that $p, \tilde{p}$ differ also by a negligible error: in particular, $p|_{z=z_s} \approx \tilde{p}_{z=z_s}$.

According to Corollary \ref{thm:bvps}, recovering $p_{z=z_s}$ entails recovering $h_s=[v_z]|_{z=z_s}$. So a procedure for computing an approximate inverse $S[c]_{s,r}^{\dagger}$ for $S[c]_{s,r}$ is 
\begin{equation}
\label{eqn:psinv}
S[c]_{s,r}^{\dagger}d_r\equiv [v_z]_{z=z_s}.
\end{equation}
where the jump in $v_z$ at $z=z_s$ is computed from knowledge of $d_r=p|_{z=z_r}$ by
\begin{itemize}
\item solving \ref{eqn:awetr} backwards in time,
\item reading off $d_s=\tilde{p}$ at $z=z_s$,
\item solving \ref{eqn:awebdry} forwards in time, and computing $[v_z]|_{z=z_s}$, or if appropriate ($\kappa$ and $\beta$ constant near $z=z_s$) simply using $\bar{v}_{z=z_s^+} = v_z^{+}=-v_z^{-}$ at $z=z_s$ (Theorem \ref{thm:bvs}).
\end{itemize}

The relations developed in the last few paragraphs may be summarized as follows:
\begin{equation}
\label{eqn:appinv}
d_r = S[c]_{s,r}h_s \,\,\Leftrightarrow \,\, h_s \approx S[c]_{s,r}^{\dagger}d_r
\end{equation}
that is, $S[c]_{s,r}^{\dagger}$ is an approximate inverse of $S[c]_{s,r}$.

Inspection of systems \ref{eqn:awebdry} and \ref{eqn:awetr} reveals that they are actually the same, except for the direction of time, and that observation leads to another useful relation. Denote by $R$ the time-reversal operator:
\[
Ru(\bx,t)=u(\bx,-t)
\]
and by {\cal R} the time reversal operator for acoustic fields:
\[
{\cal R}\left(\begin{array}{c}
p\\
\bv
\end{array}
\right) = 
\left(\begin{array}{c}
Rp\\
-R\bv
\end{array}
\right) 
\]
The sign in the definition of ${\cal R}$ is chosen so that if $(\tilde{p},\tilde{\bv})^T$ solves the system \ref{eqn:awetr}, then ${\cal R}(\tilde{p},\tilde{\bv})^T=(\hat{p},\hat{\bv})^T$ is the solution of 
\begin{eqnarray}
\label{eqn:awetrs}
\frac{\partial \hat{p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot \hat{\bv}^{\pm}, \pm(z-z_r) > 0 \nonumber \\
\frac{\partial \hat{\bv}^{\pm}}{\partial t} & = & - \beta \nabla \hat{p}^{\pm},\nonumber \\
\hat{p}^{\pm} & =& 0 \mbox{ for } t \ll 0,\nonumber\\ 
\hat{\bv}^{\pm} & = & 0 \mbox{ for } t \ll 0,\nonumber\\
\lim _{z=z_r^{\pm}}\hat{p}^{\pm} &=&  Rd_r,\nonumber\\
\hat{p} &=& \hat{p}^{\pm}, \pm(z-z_r)>0, \nonumber\\, 
\hat{\bv} &=& \hat{\bv}^{\pm}, \pm(z-z_r)>0.
\end{eqnarray}

%For a smooth orientable surface $\omega$, denote by $\Lambda[c,\omega]$ the operator that maps Dirichlet pressure data $d$ on $\omega \times \bR$ to the jump in normal velocity $[v_n]$ across $\omega \times \bR$ via solution of the system \ref{eqn:awebdry}.

 From Corollary \ref{thm:bvps} applied to the system \ref{eqn:awetrs}, $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r$. Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$. Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$. Thus the definition \ref{eqn:psinv} implies
\begin{equation}
\label{eqn:psinvrev}
S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
\end{equation}
since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.
Define $T[c]_{s,r}d_s = p|_{z=z_r}$ for the solution of \ref{eqn:awebdry}. Since ${\cal R}^2=I$, Then $\hat{p}|_{z=z_s} = T[c]_{r,s}Rd_r$. Since ${\cal R}^2=I$, ${\cal R}(\hat{p},\hat{\bv})^T=(\tilde{p},\tilde{\bv})^T$, whence $\tilde{p}|_{z=z_s}=RT[c]_{r,s}Rd_r$.

Denote by $\Lambda[c]_s$ the operator that produces the jump in normal velocity $[v_z]=\lim_{z \rightarrow z_s^+}v_z^+ - \lim_{z\rightarrow z_s^-} v_z^-$ from Dirichlet pressure data $d_s$ and causal solution of the system \ref{eqn:awebdry}. From Corollary \ref{thm:bvps}, $T[c]_{s,r} = S[c]_{s,r}\Lambda[c]_s$. Denote by $\Lambda[c]_r$ the same operator with $z_s$ replaced by $z_r$. Then $\hat{p}|_{z=z_s} = S[c]_{r,s}\Lambda[c]_rRd_r=R\tilde{p}|_{z=z_s}$. Thus the definition \ref{eqn:psinv} implies
\begin{equation}
\label{eqn:psinvrev}
S[c]_{s,r}^{\dagger}= -\Lambda[c]_sRS[c]_{r,s}R\Lambda[c]_r
\end{equation}
since $\Lambda[c]_rR=-R\Lambda[c]_r$ from the definition of ${\cal R}$.

As I will show in the example section, the approximation may be accurate enough without any further improvement. However, a closer examination shows that in fact it is possible to express the relation between $S[c]_{s,r}$ and $S[c]_{s,r}^{\dagger}$ in terms of a change of norm, that is, in terms of a preconditioner, and this relation can be used to considerably accelerate an iterative solution of the normal equation \ref{eqn:normal}.

To see this, compute the formal or $L^2$ adjoint $S[c]_{s,r}^T$, by a variant of the adjoint state method, in this case a by-product of the conservation of energy. Suppose that $\bar{h}_r$ is a function on $z=z_r$ (data traces), and 
$\bar{p},\bar{\bv}$ solve the backwards-in-time boundary value problem:
\begin{eqnarray}
\label{eqn:aweadj}
\frac{\partial \bar{p}}{\partial t} & = & \kappa (-\nabla \cdot \bar{\bv} + 
 \bar{h}_r \delta(z-z_r))\nonumber \\
\frac{\partial \bar{\bv}}{\partial t} & = & - \beta \nabla \bar{p},\nonumber \\
\bar{p} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
\bar{\bv} & = & 0 \mbox{ for } t \gg 0 
\end{eqnarray} 
Then
\[
0 = 
\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  
\frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=T}
-
\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)|_{t=0}
\]
\[
= 
\int_{0}^{T} \,dt\, \frac{d}{dt}\left(\int\, dx\,dy\,dz\, \frac{p \bar{p}}{\kappa} +  \frac{\bv \cdot \bar{\bv}}{\beta} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \frac{1}{\kappa} \frac{\partial p}{\partial t} \bar {p} +  p \frac{1}{\kappa}\frac{\partial \bar{p}}{\partial t} \right.
\]
\[
+
\left. \frac{1}{\beta} \frac{\partial \bv}{\partial t} \cdot \bar{\bv} + \frac{1}{\beta} \bv \cdot \frac{\partial \bar{\bv}}{\partial t} \right)
\]
\[
= 
\int_{0}^{T} \,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
 d_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  (- \nabla p) \cdot \bar{\bv} + \bv \cdot (-\nabla \bar{p}) \right)
\]
\[
= 
\int_{0}^{T}\,dt\, \left(\int\, dx\,dy\,dz\, \left(- \nabla \cdot \bv + 
 h_s \delta(z-z_s)\right) \bar{p} + p \left(- \nabla \cdot \bar{\bv} + 
 d_r \delta(z-z_r)\right) \right.
\]
\[
+
\left.  p (\nabla \cdot \bar{\bv}) + (\nabla \cdot \bv) \bar{p} \right)
\]
after integration by parts in the last two terms. Most of what is left cancels, leaving 
\[
0 = \int\,dt\,dx\,dy\, (h_s \bar{p})_{z=z_s} + (p  \bar{h}_r)_{z=z_r}
\]
whence
\begin{equation}
\label{eqn:sadj}
 S[c]_{s,r}^T \bar{h}_r = -\bar{p}|_{z=z_s}
\end{equation}
Compare the systems \ref{eqn:awetr} and \ref{eqn:aweadj}: they differ in that 
\begin{itemize}
\item in \ref{eqn:awetr}, $d_r$ appears as a Dirichlet condition on $z=z_r$,
\item in \ref{eqn:aweadj}, $\bar{h}_r$ appears as a distributed source on $z=z_r$.
\end{itemize}
According to Corollary \ref{thm:bvps}, $\bar{p}=\bar{p}, \bar{\bv}=\bar{\bv}$ if $\bar{h}_r = [\bar{v}_z]|_{z=z_r}$, i.e. the source is that which generates the same solution as the Dirichlet condition..

Simlarly denote by $\tilde{\Lambda}[c]_r$ the operator that produces the jump in normal velocity at $z=z_r$ from pressure data $d_r$ via {\em anticausal} solution of the system 
\begin{eqnarray}
\label{eqn:awebdrytr}
\frac{\partial {\tilde p}^{\pm}}{\partial t} & = & - \kappa \nabla \cdot {\tilde \bv}^{\pm} \mbox{ in } \pm(z-z_r) > 0, \nonumber \\
\frac{\partial {\tilde \bv}^{\pm}}{\partial t} & = & - \beta \nabla {\tilde p}^{\pm} \mbox{ in } \pm(z-z_r) > 0,\nonumber \\
{\tilde p}^{\pm} & =& 0 \mbox{ for } t \gg 0\nonumber\\ 
{\tilde \bv}^{\pm} & = & 0 \mbox{ for } t \gg 0 \nonumber\\
\lim_{z\rightarrow z_r^{\pm}}{\tilde p} & = & {\tilde d}_r 
\end{eqnarray}
%\end{thm}
 for the receiver surface $z=z_r$ (in the math literature, this is the ``Dirichlet-to-Neumann operator'', more or less). The discussion so far can be summarized as follows:
\begin{equation}
\label{eqn:wadj}
\bar{p}|_{z=z_s} = S[c]_{s,r}^T\Lambda[c]_r \bar{p}|_{z=z_r} 
\end{equation}
According to the discussion above, $\bar{p}|_{z=z_s} \approx p|_{z=z_s}$, and $h_s = \Lambda[c]_s p|_{z=z_s} = S[c]_{s,r}^{\dagger}d_r$. That is,
\begin{equation}
\label{eqn:appinv}
S[c]_{s,r}^{\dagger} = \Lambda[c]_s S[c]_{s,r}^T \Lambda[c]_r
\end{equation}
This relation can be re-written in the useful form
\begin{equation}
\label{eqn:adj}
S[c]_{s,r}^{\dagger} = W_m[c]^{-1}S[c]_{s,r}^TW_d[c]
\end{equation}
in which 
\begin{equation}
\label{eqn:weights}
W_m[c] = \Lambda[c]_s^{-1}\,\, W[c]_d = \Lambda[c]_r
\end{equation}
Since $\Lambda[c]$ is positive (semi-)definite, equation \ref{eqn:adj} exhibits $S[c]_{s,r}^{\dagger}$ as the adjoint of $S[c]_{s,r}$ with respect to weighted norms 
\begin{itemize}
\item on the source space ($h_s$): weight $W[c]_m$
\item on the data space ($d_r$): weight $W[c]_d$
\end{itemize}
Since $S[c]_{s,r}^{\dagger}S[c]_{s,r} \approx I$ as shown above, $S[c]_{s,r}$ is approximately unitary in the Hilbert spaces of source and data traces with norms defined by the weighting operators $W_m$ and $W_d$ respectively. Therefore a Krylov space method employing these norms will converge rapidly.

\cite{HouSymes:EAGE16} demonstrated a very similar preconditioner for Least Squares Migration, also for its subsurface offset extension \cite[]{HouSymes:16}, motivated by \cite{tenKroode:12}. These constructions all involve the Dirichlet-to-Neumann operator. This concept also turns up in hidden form in the work of Yu Zhang and collaborators on true amplitude migration \cite[]{YuZhang:14,TangXuZhang:13,XuWang:2012,XuZhangTang:11,Zhang:SEG09,ZhangYuSun:08,ZhangSunGray:07,ZhangBleistein:05,Bleisteinetal:05}.

>>>>

\title{Waveform Inversion via Source Extension: Overview as of Q1 2019}
\author{William. W. Symes \thanks{The Rice Inversion Project,
Department of Computational and Applied Mathematics, Rice University,
Houston TX 77251-1892 USA, email {\tt symes@caam.rice.edu}.}}

@@

%\input{prop.tex}
